Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-18 ms/main.tex
revise result 3 and spatial model

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary depending on $sign(\partial\gamma_1/\partial T_B) = sign(vb \alpha T_A+T_A-1)$. 
Therefore, increasing $T_B$ increases the boundary when $T_A > \frac{1}{1+\alpha b v}$ but when this applies we get that $\hat\gamma_1 > 1$ which guarantees cooperation will evolve from rairty since $c<1$.
When $T_A < \frac{1}{1+\alpha b v}$ increasing $T_B$ decreases the boundary.

Increasing the vertical transmission $v$ can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$ (because $\text{sign}(\partial \hat\gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.)
When $T_A<T_B$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the boundary on the cost increases making it easier for cooperation to evolve from rarity. 
Similarly, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the boundary on the cost decreases making it harder for cooperation to evolve from rarity.
We got that when $T_A>T_B$ vertical transmission is interfering cooperation to evolve by decreasing the maximum cost that the cooperates can pay and stil evolve from rairty. 
% TODO YR revise; note that without vertical transmission TA-TB determines the outcome

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if $\hat a_2 < \alpha$.
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Note that $\partial \hat a_2 / \partial v = b T_A (T_A-T_B)$. $\partial \hat a_2 / \partial v > 0$ when $T_A > T_B$ and therefore, as the vertical transmission $v$ \emph{increases} a greater value of social association is required for fixation of cooperation from rarity. %TODO YR revise
Similarly, when $T_B>T_A$ $T_A > T_B$ we get $\partial \hat a_2 / \partial v < 0$ and therefore, as $v$ \emph{decreases} a greater value of social association is required for fixation of cooperation from rarity.
\\

@@

Social association may also emerge from a population's structure.
Consider a  population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
At the beginning, each cell in the grid is randomly colonized with either a cooperator or a defector, with equal probability.
In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and with horizontal cultural transmission.
As before, $T_A$ and $T_B$ are the probabilities of transmitting phenotypes $A$ and $B$ during horizontal transmission.
At each generation, half of the individuals are randomly chosen to "initiate" interactions.
Initiators then interact with a random neighbor in a prisoners' dilemma game and a random neighbor (with replacement) for horizontal cultural transmission (as both the game and the transmission are symmetrical, the expected number of each of these interactions per individual per generation is 1).
Therefore, the effective social association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors.
On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.

The order of the interactions across the lattice at each generation is random.
After all interactions take place, and individual's fitness is determined by
$w = 1 + b \cdot n_b - c \cdot n_c$,
where $n_b$ is the number interactions of that individual with cooperative neighbors, and $n_c$ is the total number of interactions that that individual  had ($n_b \le n_c$).
Then a new generation is generated, and sites can be settled by offspring of any parent, not just neighbor parents. Thus, selection is global, rather than local, in accordance with our deterministic model.
The parent is randomly drawn with probability proportional to its fitness, divided by the average fitness of all potential parents.
Offspring then have the same phenotype as their parents.

\autoref{fig:spatial} shows that the highest cost of cooperation ($c$) that permits the evolution of cooperation in simulations of the spatial model roughly agrees with the conditions derived in Result 2.
An example for stable coexistence is shown in \autoref{fig:spatial_fluctuations}.
Simulating a spatial model with local selection (i.e. sites can only be settled by offspring of neighbor parents) had only a minor affect, eliminating the stable coexistence.

@@

%%% Figure: Spatial model - local selection
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.5]{{Ta=Tb_b=1.3_localSelection}.pdf}
    ~
    \includegraphics[scale=0.5]{{Ta=0.4_Tb=0.3-0.5_b=1.3_localSelection}.pdf} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model and local selection.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between adjacent sites, and each site had 8 neighbors.
  Selection operates locally (see \autoref{fig:spatial} for results from a model with global selection).
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in \textbf{(a)} \autoref{eq:equal_transmission} and \textbf{(b)} \autoref{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is 10,000 (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$.
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial_local_selection}
\end{figure}

% Fluctuations in spatial model
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.75]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions}.pdf}
   \caption{
    \textbf{Stable coexistence of both phenotypes in spatial model.}
    The frequancy of cooperators (green) and defactors (red) in the spatial model. 
    Both phenotypes have initial frequency of 50 precent. The figure shows that there are no fixation of nither phenotype through out the simulation.
    The frequency of both cooperators and defactors oscillates around the initial frequency. Here, population size is 10,000 (100-by-100 grid), selection operates globally, 
    benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$, horizontal transmission of cooperators $T_A = 0.4$, horizontal transmission of defactors $T_B = 0.435$ and cost $c=0.02$.
    }
    \label{fig:spatial_fluctuations}
  \end{figure}
  %TODO: edit this caption. DC

>>>>

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection. In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.

Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation, $T_A$, increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), facilitating the evolution of cooperation.
However, increasing the horizontal transmission of defection, $T_B$, can increase or decrease the cost boundary, but it increases the cost boundary when it is already above one ($c<1<\hat\gamma_1$): $\partial\gamma_1/\partial T_B$ is positive when $T_A > \frac{1}{1+\alpha b v}$, which gives $\hat\gamma_1>1/v$. 
Therefore, increasing $T_B$ decreases the cost boundary and limits the evolution of cooperation, but only if $T_A < \frac{1}{1+\alpha b v}$.

Increasing the vertical transmission $v$ can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$, because $\text{sign}(\partial \hat\gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.
When $T_A<T_B$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the cost boundary increases, making it easier for cooperation to evolve from rarity. 
In contrast, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the cost boundary decreases making it harder for cooperation to evolve from rarity.

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if $\hat a_2 < \alpha$.
\end{corollary}
\autoref{fig:result3}cd illustrates this condition.
Interestingly, the effect of vertical transmission on this condition depends on the horizontal transmission bias, $\partial \hat a_2 / \partial v = b T_A (T_A-T_B)$. 
That is, with horizontal bias for cooperation, $T_A>T_B$, the boundary increases with the rate of vertical transmission, impeding the evolution of cooperation.
In contrast, with horizontal bias for defection, $T_A<T_B$, the boundary decreases with the rate of vertical transmission, facilitating the evolution of cooperation. 
These effects of vertical transmission are demonstrated in \autoref{fig:result3}cd.
\\

@@

Social association may also emerge from a population's structure.
Consider a  population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized with either a cooperator or a defector, with equal probability.
%In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and in horizontal cultural transmission (\autoref{fig:horizontal}).
%As before, $T_A$ and $T_B$ are the probabilities of transmitting phenotypes $A$ and $B$ during horizontal transmission.
In each generation, half of the individuals are randomly chosen to "initiate" interactions.
Initiators then interact with a random neighbor (i.e. individual in an adjacent site) in a prisoners' dilemma game (\autoref{table:prisoner_payoff}) and a random neighbor (with replacement) in horizontal cultural transmission (\autoref{fig:horizontal}).
The expected number of each of these interactions per individual per generation is one.
The effective social association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors.
On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probability to successfully transmitting phenotypes $A$ and $B$ during horizontal transmission, respectively.

The order the interactions across the grid at each generation is random.
After all interactions take place, individual fitness is determined by
$w = 1 + b \cdot n_b - c \cdot n_c$,
where $n_b$ is the number interactions of that individual with cooperative neighbors, and $n_c$ is the total number of interactions that that individual  had ($n_b \le n_c$).
Then a new generation is generated, and sites can be settled by offspring of any parent, not just neighbor parents. Thus, selection is global, rather than local, in accordance with our deterministic model.
The parent is randomly drawn with probability proportional to its fitness, divided by the average fitness of all potential parents.
Offspring then have the same phenotype as their parents ($v=1$).

\autoref{fig:spatial} shows that the highest cost of cooperation ($c$) that permits the evolution of cooperation in simulations of the spatial model agrees with the conditions derived in Result~2.
An example for stable coexistence is shown in \autoref{fig:spatial_fluctuations}.
Simulating a spatial model with local selection (i.e. sites can only be settled by offspring of neighbor parents) had only a minor effect on the agreement with Result~2 (\autoref{fig:spatial_local_selection}).

@@

%%% Figure: Spatial model - local selection
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.5]{{Ta=Tb_b=1.3_localSelection}.pdf}
    ~
    \includegraphics[scale=0.5]{{Ta=0.4_Tb=0.3-0.5_b=1.3_localSelection}.pdf} \\
  \caption{
  \textbf{Evolution of cooperation in a spatial model with local selection.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the cost of cooperation ($c$) on the y-axis, and the symmetric horizontal transmission rate ($T=T_A=T_B$) on the x-axis of the left panel, or the transmission bias $T_A-T_B$ on the x-axis of the right panel.
  The population evolves on a 100-by-100 grid.
  Cooperation and horizontal transmission are both local between adjacent sites, and each site had 8 neighbors.
  Selection operates locally (see \autoref{fig:spatial} for results from a model with global selection).
  The black curves represent the cost boundaries for the evolution of cooperation in a well-mixed population with social association where $\alpha=1/8$ in \textbf{(a)} \autoref{eq:equal_transmission} and \textbf{(b)} \autoref{eq:cost_boundaries}.
  Simulations were stopped at generation 10,000 or if one of the phenotypes fixed and 50 simulations were executed for each parameter set.
  Here, population size is 10,000 (100-by-100 grid), benefit of cooperation, $b=1.3$, perfect vertical transmission $v=1$.
  \textbf{(a)} Symmetric horizontal transmission, $T=T_A=T_B$.
  \textbf{(b)} Horizontal transmission rates $T_A=0.4$, $0.3<T_B<0.5$.
  }
  \label{fig:spatial_local_selection}
\end{figure}

% Fluctuations in spatial model
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.75]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions}.pdf}
   \caption{
    \textbf{Stable coexistence of both phenotypes in a spatial model.}
    The frequency of cooperators (green) and defectors (red) in a spatial model. 
    Both phenotypes start at 50\% frequency.
    The figure shows that neither phenotype is fixed throughout the simulation, maintaining a stochastic coexistence.
    Here, population size is 10,000 (100-by-100 grid), selection operates globally, 
    benefit and cost of cooperation $b=1.3$ and $c=0.02$, perfect vertical transmission $v=1$, horizontal transmission rates $T_A = 0.4 < T_B = 0.435$.
    }
    \label{fig:spatial_fluctuations}
  \end{figure}
  %TODO: edit this caption. DC
